---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# morphospace

<!-- badges: start -->
<!-- badges: end -->

A package for people who can't visualize morphospaces good and wanna learn to do other geometric morphometrics stuff good too.

## Installation

You can install the development version of morphospace from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("millacarmona/morphospace")
```

Load tail data and extract shapes, centroid sizes, classification of sex and species,
links between landmarks, and phylogenetic tree

```{r}
library(morphospace)

data("tails")
shapes <- tails$shapes
species <- tails$data$species
sizes <- tails$sizes
sex <- tails$data$sex
links <- tails$links
tree <- tails$tree
```

Morphometric variation is assumed to be already free of differences of orientation, position and scale. This standardization can be readily performed using the capabilities of `R` packages such as `Morpho`, `geomorph`, and `Momocs`.

```{r pressure, echo = FALSE}
# Inspect shapes
pile_shapes(shapes, links = links)
```



```{r cars}
summary(cars)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots, for example:

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
